---
title: "Introduction to Regression Analysis"
subtitle: "Social Data Institute, Day 2"
author: "Michal Ovádek"
date: ""
format: revealjs
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)

```

## Today

- revisit statistical foundations
- discuss key ideas in causal inference
- work with more data

## Units and variables 

There are 2 organising features of any data that we study: units and variables.

**Units** ($i \in 1,...,N$)

    -   The objects that we are studying
    -   Usually these are the rows of the dataset
    -   E.g. individuals; countries; companies; Members of Parliament; etc
    -   We usually use $i$ to indicate a unit, and $N$ to mean the total number of units

**Variables**

    -   Measurements of characteristics that vary across units
    -   Usually these are the columns of the dataset
    -   E.g. age; income; vote choice; profit/loss; GDP; etc
    
## Dependent and independent variables

An important conceptual distinction between types of variable:

-   **Dependent variable (**$Y$)

    -   Variable to be explained
    -   Also called the outcome or response variable

-   **Independent variables (**$X$)

    -   Determinant(s) of the dependent variable
    -   Also called the explanatory or predictor variables
    -   Sometimes (somewhat confusingly) expressed as $T$ or $D$

## Levels of measurement

### Continuous/Interval 

-   Values indicate precise differences between categories
-   Differences (intervals) have the same meaning anywhere on the scale
-   E.g age

---

### Categorical/Nominal 

-   Values indicate different, mutually exclusive categories
-   No relative information in the categories
-   E.g political parties, gender

---

### Ordinal 

- Values indicate relative differences between categories
- Imply a ranking, but difference between categories may be unknown
- E.g. educational achievement

Determining the correct level of measurement is important for making decisions about how to analyse your data.

## Sums and Sigma notation

-   $N$ is the number of units or the **sample size**

-   If $N = 100$, we have 100 measurements of each variable ($Y_1, Y_2, Y_3,...,Y_N$)

-   We will often want to refer to the **sum** of a variable, e.g. $Y_1 + Y_2 + Y_3 + ... + Y_N$. But this gets cumbersome if $N$ is large!

. . .

-   Instead, we will often use **Sigma notation**: $$\sum_{i = 1}^N Y_i = Y_1 + Y_2 + Y_3 + ... + Y_N$$  where $\sum_{i=1}^N Y_i$ means "sum up all instances of $Y$ starting from 1 and ending at N".

## Measures of central tendency

To compare observations, we need some way of summarising their characteristics. The most common summaries for most variables are those that measure the **central tendency** of the variable.

## Central Tendency

The value of a "typical" observation, or the value of the observation at the center of a variable's distribution.

We will consider three measures of central tendency:

1.  Mean
2.  Median
3.  Mode

## Mean

The mean is the "average" or "expected" value of a variable.

It is denoted $\bar{Y}$ or $\bar{X}$, which can be read as "Y bar" or "X bar".

$$\bar{Y} = \frac{\sum_{i = 1}^N Y_i}{N} = \frac{1}{N}\sum_{i=1}^N Y_i$$

. . .

- I.e. we add up the values of $Y$ and divide by the sample size.

## Median

The median is the value of a variable that divides the data into two groups such that there are an equal number above and below.

$$
Median =
\begin{cases}
x_{((N+1)/2)} & \text{when N is odd}\\
\frac{1}{2}\left( x_{(N/2)} + x_{(N/2+1)}\right) & \text{when N is even}
\end{cases}
$$

where $x_{i}$ is the $i^{th}$ smallest value of variable $x$.

. . .

- I.e. the median is the middle value when the total number of observations is odd, and the average of the two middle values when the total number of observations is even

## Mode

The mode is simply the most common value of a variable.

## Mean, Median or Mode?

Which measure we use depends on the level of measurement:

-   The **mean** is most appropriate for *continuous* variables
-   The **median** is most appropriate for *ordinal* variables
-   The **mode** is most appropriate for *categorical* variables

Let's compute them for the British Election Study (BES) data.

## Beyond the Central Tendency

Central tendency is not the only potential summary we might care about.

We may be interested in how the data is *distributed* *around* the measures of central tendency.

## Quantiles

- The `summary()` command provided the 0th, 25th, 50th (median), 75th and 100th percentile.
    + The Xth "percentile" (0-100) is the value below which X percent of observations in a group of observations fall.
    + The Xth "quantile" (0-1) is the value below which X proportion of observations in a group of observations fall.
- You can request any specific quantile of the distribution you want
    + The 5th, 50th and 95th percentiles are the 0.05, 0.5 and 0.95 quantiles, respectively.
    
## Summarising Dispersion

How do we summarise *variation* or *dispersion* in a data set?

How do we describe the difference between these distributions, all of which have the same mean/median?

- 20, 20, 20, 20, 20
- 15, 18, 20, 22, 25
- 10, 15, 20, 25, 30
- 0, 10, 20, 30, 40

## Range

One measure of dispersion is the range, the difference between the 0th and 100th percentile (or equivalently, the 0.00 and 1.00 quantile)

- **20**, 20, 20, 20, **20** $\rightarrow$ 0
- **15**, 18, 20, 22, **25** $\rightarrow$ 10
- **10**, 15, 20, 25, **30** $\rightarrow$ 20
- **0**, 10, 20, 30, **40** $\rightarrow$ 40

This turns out to be a poor choice in practice, because it is very sensitive to changes in a single extreme value.

## Interquartile Range

A better measure of dispersion is the interquartile range, the difference between the 25th and 75th percentile (or equivalently, the 0.25th and 0.75th quantile)

- 20, **20**, 20, **20**, 20 $\rightarrow$ 0
- 15, **18**, 20, **22**, 25 $\rightarrow$ 4
- 10, **15**, 20, **25**, 30 $\rightarrow$ 10
- 0, **10**, 20, **30**, 40 $\rightarrow$ 20

## Standard deviation

While the interquartile range is easy to understand, far more frequently we use the *standard deviation* as a measure of dispersion.
$$s = \sqrt{\frac{1}{N-1} \sum_{i=1}^N (x_i - \overline{x})^2}$$
---


**Building the standard deviation:** Start with the difference between each observation and the mean:

$$(x_i - \overline{x})$$

---


**Building the standard deviation:** Square these so that bigger deviations from the mean (whether positive or negative) are more positive numbers

$$(x_i - \overline{x})^2$$

---


**Building the standard deviation:** Add them all up

$$\sum_{i=1}^N (x_i - \overline{x})^2$$

---

**Building the standard deviation:** Divide by the number of things you added up (minus one, for technical reasons that need not concern us here).

$$\frac{1}{N-1} \sum_{i=1}^N (x_i - \overline{x})^2$$

---

**Building the standard deviation:** Finally, take a square root to "undo" the square, so that the units of the standard deviation are on the original scale of $x$.

$$\sqrt{\frac{1}{N-1} \sum_{i=1}^N (x_i - \overline{x})^2}$$

## Interpreting the standard deviation

- Larger standard deviations mean that the data are more dispersed / variable.
- Smaller standard deviations mean that the data are less dispersed / variable.
     + A standard deviation of 0 occurs only if all the observations have the same value.

Let's now implement it in R on the BES data.

## z-scores

Sometimes it is useful to "standardize" variables in terms of their standard deviation.

- This is typically done by calculating "z-scores"

- This preserves the relative values of $x$, but the new variable $z$ then has a mean of exactly $0$ and a standard deviation of exactly $1$
    
$$ z_i = \frac{x_i - mean(x)}{sd(x)}$$

## Dispersion in data

```{r, echo = FALSE}

load("data/bes.Rda")

d1 <- rnorm(100, mean = 0, sd = 1)
d2 <- rnorm(100, mean = 0, sd = 5)
d3 <- rnorm(100, mean = 0, sd = 10)

par(mfrow = c(3,1))
plot(y = 1:100, x = d1, xlim = c(-25, 25), xlab = NA, ylab = NA)
plot(y = 1:100, x = d2, xlim = c(-25, 25), xlab = NA, ylab = NA)
plot(y = 1:100, x = d3, xlim = c(-25, 25), xlab = NA, ylab = NA)

```

## Histograms

We will often use histograms to get an understanding of the distribution of a given variable.

```{r, echo=TRUE}
# create data
random_sample <- data.frame(var = rnorm(1000, mean = 0, sd = 10))

# plot
random_sample |> 
  ggplot(aes(x = var)) +
  geom_histogram()
```

## Boxplots

Boxplots are another common way to visualize data.

```{r, echo=TRUE}
bes |> 
  ggplot(aes(y = age, x = turnout)) +
  geom_boxplot()
```

## Correlation

The most commonly used bivariate descriptive statistic is the *correlation coefficient*:

$$
      \rho =  \frac{1}{n-1} \sum_{i=1}^n \left(\frac{x_i - \bar{x}}{S_x} \times \frac{y_i - \bar{y}}{S_y}\right)
$$

The correlation coefficient describes the strength of association between two continuous variables.

## Examples of correlation coefficients

```{r,fig.width=12,fig.height=8}

par(mfrow=c(2,3),mar=c(5,5,5,5))

corr_to_plot <- c(-1,0,1,0.5,0.75,0.9)
for (i in 1:6){
  tmp <- MASS::mvrnorm(500,c(0,0),matrix(c(1,corr_to_plot[i],corr_to_plot[i],1),2,2))
  plot(tmp[,1],tmp[,2],xlab="",ylab="",axes=FALSE,main=paste(expression(rho),"=",corr_to_plot[i]),col=rgb(0,0,0,0.5),pch=16)
}

```


## What is a large correlation?

- What can be considered a 'large' correlation?

    + $\rho = 1$ is a perfect positive association
    + $\rho = 0$ is no association
    + $\rho = -1$ is a perfect negative association.
    
- There is no fixed translation of numbers to words that applies generally

    + In some contexts $\rho=0.5$ is "strong", in others it is "moderate", in others it might be "weak". Do we have a sense of a baseline for comparison?

# Causality and counterfactuals

## Why causality?


Cause-and-effect relationships are at the heart of some of the most interesting (social) science theories.

. . .

We can often express important research topics as a simple question: Does *X* cause *Y*?

. . .

-   Does *economic development* lead to *democracy*?
-   Does the *race of a politician* affect their *chances of being elected*?
-   Does *immigration* increase support for *right-wing parties*?
-   Does *health insurance* lead to better *health*?

## Example

**Does health insurance improve health outcomes?**

A critical issue of public policy is whether and how governments should provide health care to their citizens. But is there a *causal* effect of
health insurance on actual levels of health?

-   **Y (Dependent variable, or "outcome")**: *Health*

    -   What is the self-assessed health of an individual?

-   **X (Independent variable, or "treatment")**: *Health insurance*

    -   Does the individual have health insurance, or not?

## Counterfactuals

Whenever a person makes a causal statement, they are contrasting (often implicitly)...

1. what they observe (something *factual*) with 
2. what they believe they would observe if a key condition was different (something *counterfactual*).

-   "Austerity caused Brexit" ([Fetzer, 2019](https://warwick.ac.uk/fac/soc/economics/research/centres/cage/manage/publications/381-2018_fetzer.pdf))

    -   Observation: Austerity policy in the UK and a vote to leave the EU
    -   Belief: Without austerity, smaller Leave vote

---

-   "Oil wealth inhibits democracy" ([Ross, 2013](https://press.princeton.edu/titles/9686.html))

    -   Observation: Oil-rich countries tend to be less democratic
    -   Belief: If countries had less oil they would be more democratic
    
## Unseen Counterfactuals 

- In our example: Health insurance $\rightarrow$ health

    -   Observation: People with health insurance are healthier
    -   Belief about counterfactual: If people had no health insurance, they would be less healthy \pause
    - But we cannot observe the **same** person with *and* without insurance at once 
    
. . .

Problem: We can't observe counterfactual outcomes! Causal inference requires *estimating* counterfactuals for comparison to realised outcomes.

## Treatment and outcome

We will think about causal relationships in terms of effects of **treatments** on outcomes.

. . .

-   Treatment: Where change originates
-   Outcome: What is affected by change

. . . 

We will focus on binary treatment variables:

-   $X_i$ is 1 if observation $i$ is treated
-   $X_i$ is 0 if observation $i$ is not treated

We will focus on continuous/interval outcome variables

## Treatment and outcome (example)

In our example:

-   $X_i$ is 1 if individual $i$ has health insurance
-   $X_i$ is 0 if individual $i$ does not have health insurance
-   $Y_i$ is individual $i$'s health

. . .

Keep in mind: We can't observe counterfactual outcomes. Causal inference requires *estimating* counterfactuals for comparison to realised outcomes.

## Potential outcomes and causal effects

We can *define* the causal effect of an independent/treatment variable $X$ on an outcome variable $Y$ by considering the **potential** outcomes of $Y$.

---

### Potential outcomes

The potential outcomes of $Y$ are the values of $Y$ that would be realised for different values of $X$. e.g.

-   $Y_i(1)$ = the value $Y_i$ *would have been* if $X_i$ was equal to 1
-   $Y_i(0)$ = the value $Y_i$ *would have been* if $X_i$ was equal to 0


### Treatment effect

For any given individual, if we could observe both potential outcomes, the treatment effect of X on Y for that individual can be calculated as $$Y_i(1) - Y_i(0)$$

## Potential outcomes and causal effects (example)

What are the potential outcomes in our health insurance example?

-   $Y_i(1) =$ The health individual $i$ would have if the individual had health insurance
-   $Y_i(0) =$ The health individual $i$ would have if the individual did not have health insurance

---

What are the treatment effects?

-   If $Y_i(1) > Y_i(0)$ then insurance improves health
-   If $Y_i(1) < Y_i(0)$ then insurance worsens health
-   If $Y_i(1) = Y_i(0)$ then insurance has no effect on health

## Potential outcomes and causal effects (example)

-   $X_i = 1$ if the individual is insured, and $X_i=0$ if uninsured
-   $Y_i(1)$ is the health of the individual if they were insured
-   $Y_i(0)$ is the health of the individual if they were uninsured
-   The treatment effect for an individual is $Y_i(1) - Y_i(0)$

## Potential outcomes and causal effects (example)

| Individual | $X_i$ |      $Y_i(1)$      |      $Y_i(0)$      |  Treatment effect  |
|:----------:|:-----:|:---------------:|:--------------:|:----------------------:|
|     1      |   1   | 5  | 3  | 2  |
|     2      |   1   | 5  | 4  | 1  |
|     3      |   0   | 3  | 3  | 0 |
|     4      |   0   | 4 | 3 | 1 |

Average treatment effect (ATE) = $\frac{2+1+0+1}{4} = \frac{4}{4} = 1$

## The Fundamental Problem of Causal Inference

**But** we cannot observe *both* potential outcomes for any individual!

| Individual | $X_i$ |      $Y_i(1)$      |      $Y_i(0)$      |  Treatment effect  |
|:----------:|:-----:|:---------------:|:--------------:|:----------------------:|
|     1      |   1   | 5  | ?  | ?  |
|     2      |   1   | 5  | ?  | ?  |
|     3      |   0   | ?  | 3  | ? |
|     4      |   0   | ? | 3 | ? |

Average treatment effect (ATE) = ?


## The Fundamental Problem of Causal Inference

We only ever observe *one* potential outcome for a given individual, and our observed outcome depends on the status of our explanatory variable. We can never directly observe individual causal effects.

### Consequences

-   We cannot compute causal effects for individuals.
-   We have to *estimate* counterfactuals for comparison to realised outcomes.

## Quantity of Interest

We want to estimate the **average treatment effect (ATE)**: $$\frac{1}{N}\sum_{i=1}^N \{Y_i(1) - Y_i(0)\}$$ But we can't observe $Y_i(1)$ *and*
$Y_i(0)$ for any given unit! 

. . . 

One alternative is to use the **difference-in-means** to *estimate* the ATE: $$\bar{Y}_{X=1} - \bar{Y}_{X=0}$$ where $\bar{Y}_{X=1}$ and
$\bar{Y}_{X=0}$ are the average *observed* outcomes for treated and control units, respectively.

## Confounding

The **difference-in-means** estimator is only unbiased if there is no confounding.

Example 1: Causal effect of finance degree on income tax policy preferences

-   Finding: People who study finance are more likely to prefer lower taxes
-   Confounding: People who want to study finance might prefer lower income tax to begin with
-   Direction of bias: Positive, we overstate the effect of a finance degree

---

Example 2: Causal effect of aspirin on headache symptoms

-   Finding: People who take aspirin report worse headache symptoms
-   Confounding: People who took aspirin had headaches to begin with
-   Direction of bias: Negative, we understate the effectiveness of aspirin

## **Does health insurance improve health outcomes?**

The National Health Interview Survey (NHIS) is an annual survey of the US population that asks questions about health and health insurance. We are
interested in two questions in particular:

-   Y (Dependent variable): *health*

    -   "Would you say your health in general is excellent (5), very good (4), good (3), fair (2), or poor (1)?"

-   X (Independent variable): *insured*

    -   "Do you have health insurance?"
    
## NHIS data

```{r echo=TRUE}
load("data/nhis.Rda")

summary(nhis)

# now estimate the ATE of insurance in R
```

## NHIS data

1.  Can we interpret the difference in means here as causal?

2.  How might we assess the possibility of confounding bias here?

## Checking "balance"

One implication of confounding:

- Treatment and control units will be different with respect to characteristics other than the treatment.

- We can check this by evaluating the degree of **balance** for treated and control units on different pre-treatment variables.

Let's calculate balance in R.


## Randomised experiments and observational studies


::: columns
::: {.column width="50%"}

**Randomised experiments**

- Units are randomly assigned to different X values
- Researchers directly intervene in the world they study
- Gold standard for causal inference

:::

::: {.column width="50%"}

**Observational studies**

- Units are assigned to X values "by nature"
- Researchers observe, but do not intervene in, the world
- Very commonly used in social science research

:::
:::


## Randomisation and selection bias

We saw previously that, in general, the difference-in-means will not be an unbiased estimate of the ATE because of confounding.

Question: How can we avoid this problem?

Answer: Conduct a randomised experiment!

---

### Key intuition

Randomisation of treatment doesn't eliminate differences between individuals, but ensures that the mix of individuals being compared is the same on average.

### Consequence

We cannot use randomisation to calculate individual causal effects, but we can use it to estimate average causal effects.

## Randomisation and selection bias

When the treatment, $X$, is randomly assigned to units... 

-   ... treated and untreated groups will be similar, on average, in terms of all characteristics (both observed and unobserved)
-   ... the only *systematic* difference between the two will be the receipt of treatment
-   ... the average outcome for controls will be similar to what *would have happened* for the treatment group if they had not been treated

## Randomisation and selection bias

Randomisation of the treatment makes the difference in group means an **unbiased** estimator of the true ATE.

- It 'solves' the problem of confounding by not letting units *select* into treatment

- *On average*, you can expect a randomised experiment to get the right answer

- This does not guarantee that the answer you get from any particular randomisation will be exactly correct


## Example: Experimental data

The RAND Health Insurance Experiment (RAND) was an experiment conducted between 1974 and 1982 in the US. In this experiment, researchers *randomly
allocated* individuals to receive health insurance.

-   Y (Dependent variable): *health*

    -   "Would you say your health in general is excellent (5), very good (4), good (3), fair (2), or poor (1)?"

-   X (Independent variable): *insured*

    -   Was the participant *randomly allocated* to receive health insurance?

## RAND data

```{r echo=TRUE}
load("data/rand.Rda")

summary(rand)

# now estimate the ATE of insurance in R
```

## External and internal validity

**Internal validity:** Can we interpret estimates in a study as causal? 

-   Randomised experiments tend to have high internal validity

    + $\rightarrow$ randomisation makes treatment and control groups similar on average
    
-   Observational studies tend to have lower internal validity

    + $\rightarrow$ pre-treatment variables may differ between treatment and control

## External and internal validity


**External validity:** How generalizable are the conclusions of a study? 

-   Randomised experiments tend to have lower external validity 

    + $\rightarrow$ difficult to conduct on representative samples 
    
-   Observational studies tend to have higher external validity 

    + $\rightarrow$ easier to use representative samples or the population itself

## External vs internal validity

::: columns
::: {.column width="50%"}

### Internal Validity

Credibility of the results as *causal*.
:::
::: {.column width="50%"}

### External Validity

Credibility of the results as *generalisable*.
:::
:::

